Formalizing Stack Safety as a Security Property

### 2023-02-04 submission5/core.tex
no message

{\em Functions} in high-level languages, and related abstractions such
as subroutines, procedures, methods, etc., are units of computation
that call one another to modularly realize larger computations.
%
At a low level, each function activation manages its own data---local
variables, etc.---as well as information about the \emph{caller} to
which it must return.
%
The \emph{call stack} is the fundamental data structure used to
implement functions, aided by an ABI that defines how registers are shared
between activations.

@@

We define stack safety properties in
terms of traces of {\em observable events} rather than internal machine states.
This is particularly important for
specifying lazy enforcement, which works because the deferred check catches problems
before they can affect the trace.
Events might include certain function calls (i.e., system calls performing I/O)
or writes to special addresses representing memory-mapped regions, etc.
%The nature of events is a parameter of the framework, and can be specialized
%to any notion of observable behavior.

Finally, we define our high-level security properties---integrity, confidentiality,
and well-bracketed control flow---in terms of the security classes of state
elements. These take the form of predicates on states and contexts that use variants,
assertions about future return states, and comparison of event traces to capture the
desired behavior without reference to internal implementation details.
We introduce these properties by example in \cref{sec:example}. In the remainder
of this section we describe the underlying framework in more detail.

Our security semantics has similarities to the overlay semantics proposed
by Skorstengaard et al.~\cite{SkorstengaardSTKJFP}, but it does not restrict
the behavior of the underlying machine in any way. Rather, it tracks additional context
about the history of security-relevant operations, which inform the criterion a machine
must satisfy in order to correctly obey the properties.\apt{Sharpen this comparison?}

@@

\paragraph*{Machine model}
We assume a conventional processor ISA (e.g. RISC-V, x86-64, etc.), with registers including a program counter
and stack pointer, and a standard ABI.
We make no particular assumptions about the provenance of the machine code; in particular,
we do not assume the use of any particular compiler.
The machine may possibly be enhanced with security
enforcement mechanisms such as hardware tags~\cite{pump_hasp2014,Gollapudi+23} or capabilities~\cite{Woodruff+14}. If so,
we assume that the behavior of these mechanisms is incorporated into the basic
step semantics of the machine, but we distinguish the \emph{payload} part of
each state element---the part relevant to the behavior of the basic machine---from
any additional enforcement information (such as a tag).

\paragraph*{Security Semantics and Property Structure}
A security semantics extends a machine
with additional context about the identity of current and pending
functions (which act as security principals) and about the registers and memory they require
to be secure. This added context is purely notional;
it does not affect the real machine. The security context
evolves dynamically through the execution of security-relevant operations,
which include events like calls, returns, and frame manipulations.
Our security properties are phrased in terms of this context, often as predicates
on future states, e.g. of the form ``when control returns to the current function...'',
or as relations on traces of future execution (hyper-properties).

The security-relevant operations typically correspond to underling machine instructions,
but the correspondence may not be obvious just by inspecting the machine code.
For example, in the tagged RISC-V machine we use in our examples and tests,
calls and returns are conventionally performed by {\tt jal} (``jump-and-link'')
and {\tt jalr} (``jump-and-link-register'') instructions, respectively, but these
instructions might also be used for other things. An instantiation of an operation
might require some external mechanism for distinguishing which instructions are
security-relevant. The most obvious technique, which we use in our examples, is to add labels to
selected instructions, but the theory does not depend on this choice.
Whatever the mechanism for labeling transitions, we end up with an annotated version of the
machine transition function \(\mach \xrightarrow{\bar{\psi}, \obs} \mach'\), where \(\obs\) is
an event and \(\bar{\psi}\) is a list of security-relevant operations.

We then lift this into a transition between pairs of machine states and contexts,
by applying a set of operation-specific rules to transform each state and context to
a new context, in parallel with the ordinary transition on states.
The most important rules describe call and return operations.
A call pushes a new activation view onto the context stack and changes the class of the
caller's data to protect it from the new callee; a return reverses these steps.
Other operations can serve to signal how parts of the stack frame are being used to store
or share data, and their corresponding rules alter the classes of different
state elements accordingly.

Exactly which operations and rules are needed depends on
what code features we wish to support.
The set of security-relevant operations (\(\Psi\)) covered in this paper is given in
\cref{tab:psi}. Those that we explore in detail in \cref{sec:example,sec:formal}
are shown in white: calls, returns, and the allocation and deallocation of local memory.
In \cref{sec:extensions}, we cover additional operations and and parameters
to support various forms of memory sharing \ifexceptions , exceptions, \fi
and tail-call elimination; these are shown in light gray. These are also
the operations we test in \cref{sec:testing}.
We give a proof-of-concept formalization of a more sophisticated, heap-like sharing model
based on pointer provenance in \cref{app:ptr}, represented by the dark gray
operations, but we do not test this.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l | l |}
    \hline
    \(\psi \in \Psi\) & Parameters \\
    \hline
    \example \(\mathbf{call}\) & target address, argument registers \\
    \testing & stack arguments (base, offset \& size) \\
    \example \(\mathbf{return}\) & \\
    \example \(\mathbf{alloc}\) & offset \& size \\
    \testing & public flag \\
    \example \(\mathbf{dealloc}\) & offset \& size \\
    \testing \(\mathbf{tail call}\) & (same as for \(\mathbf{call}\)) \\
    \hline
    \multicolumn{2}{|c|}{{\it Pointer provenance operations}} \\
    \hline
    \theory \(\mathbf{promote}\) & register, offset \& size \\
    \theory \(\mathbf{propagate}\) & source register/address \\
    \theory & destination register/address \\
    \theory \(\mathbf{clear}\) & target register/address \\
    \hline
  \end{tabular}
\end{center}
\caption{Security-relevant operations and their parameters.}
  \label{tab:psi}
\end{table}

>>>>

{\em Functions} in high-level languages, and related abstractions such
as subroutines, procedures, methods, etc., are units of computation
that call one another to modularly realize larger computations.
%
At a low level, each function activation manages its own data---local
variables, etc.---as well as information about the \emph{caller} to
which it must return.
%
The \emph{call stack} is the fundamental data structure used to
implement functions, aided by an ABI\bcp{will all CSF readers know
  what an ABI is?} that defines how registers are shared
between activations.

@@

We define stack safety properties in
terms of traces of {\em observable events} rather than internal machine states.
This is particularly important for
specifying lazy enforcement, which catches bad behaviors after they occur but
before they can affect the trace.
Observable events might include specific function calls (e.g., system
calls that perform I/O), writes to special addresses representing
memory-mapped regions, etc.
%The nature of events is a parameter of the framework, and can be specialized
%to any notion of observable behavior.

Finally, we define high-level security properties---integrity, confidentiality,
and well-bracketed control flow---in terms of the security classes of state
elements. These take the form of predicates on states and contexts
that use variants\bcp{undefined},
assertions about future return states, and comparison of event traces to capture the
desired behavior without reference to internal implementation details.
We introduce these properties by example in \cref{sec:example} and
formally in \cref{sec:formal}. In the remainder
of this section we describe the underlying framework in more detail.

Our security semantics has similarities to the overlay semantics
proposed by Skorstengaard et al.~\cite{SkorstengaardSTKJFP}, but it
does not restrict the behavior of the underlying machine in any
way\bcp{Will people not familiar with Skorstengaard understand
  that?}. Rather, it tracks additional context about the history of
security-relevant operations, which informs the criterion a machine
must satisfy in order to correctly obey the properties.\apt{Sharpen
  this comparison?}\bcp{+1.  Or maybe, in the interest of time and to
  keeping this section flowing, move it to related work.}

@@

\paragraph*{Machine model}
We assume a conventional ISA (e.g. RISC-V, x86-64, etc.), with registers including a program counter
and stack pointer, and a standard ABI.
We make no particular assumptions about the provenance of the machine code; in particular,
we do not assume the use of any particular compiler.
The machine may possibly be enhanced with security
enforcement mechanisms such as hardware tags~\cite{pump_hasp2014,Gollapudi+23} or capabilities~\cite{Woodruff+14}. If so,
we assume that the behavior of these mechanisms is incorporated into the basic
step semantics of the machine, but we distinguish the \emph{payload} part of
each state element---the part relevant to the behavior of the basic machine---from
any additional enforcement information (such as a tag).\bcp{Why?}

\paragraph*{Security Semantics and Property Structure}
A security semantics extends a machine
with additional context about the identities of current and pending
functions (which act as security principals) and about the registers and memory they require
to be secure. This added context is purely notional;
it does not affect the behavior of the real machine. The security context
evolves dynamically through the execution of security-relevant operations,
which include events like calls, returns, and frame manipulation.
Our security properties are phrased in terms of this context, often as predicates
on future states (``when control returns to the current function...'')
or as relations on traces of future execution
(hyper-properties).\bcp{This is the first we've mentioned
  hyperproperties.  CSF readers may be surprised.}

Security-relevant operations typically correspond to
\proposeadd{single?} underlying machine instructions,
but the correspondence may not be obvious just by inspecting the machine code.
For example, in the tagged RISC-V architecture that we use in our
examples and tests,
calls and returns are conventionally performed using the {\tt jal}
(``jump-and-link'')
and {\tt jalr} (``jump-and-link-register'') instructions, but these
instructions might also be used for other things. An instantiation of
an operation
might require some external mechanism for distinguishing which instructions are
security-relevant.\bcp{Huh?  What's an instantiation of an operation?} The most obvious technique, which we use in our examples, is to add labels to
selected instructions, but the theory does not depend on this
choice.\bcp{What could one do instead?}

Whatever the mechanism for labeling transitions, we end up with an annotated version of the
machine transition function: we write \(\mach \xrightarrow{\bar{\psi},
  \obs} \mach'\), where \(\mach\) and \(\mach\) are machine states, \(\obs\) is
an event, and \(\bar{\psi}\) is a list of security-relevant operations.
%
We then lift this into a transition between pairs of machine states and contexts,
by applying a set of operation-specific rules to transform each state and context to
a new context, in parallel with the ordinary transition on states.
The most important rules describe call and return operations.
A call pushes a new activation view\bcp{it was just called ``view'' above} onto the context stack and changes the class of the
caller's data to protect it from the new callee; a return reverses these steps.
Other operations can serve to signal how parts of the stack frame are being used to store
or share data, and their corresponding rules alter the classes of different
state elements accordingly.

Exactly which operations and rules are needed depends on
what code features we wish to support.
The set of security-relevant operations (\(\Psi\)) covered in this paper is given in
\cref{tab:psi}.\bcp{I was confused by the table because I looked at
  the figure before I read that sentence!} Those that we explore in detail in \cref{sec:example,sec:formal}
are shown in white: calls, returns, and the allocation and deallocation of local memory.
In \cref{sec:extensions}, we cover additional operations and and parameters
to support various forms of memory sharing \ifexceptions , exceptions, \fi
and tail-call elimination; these are shown in light gray. These are also
the operations we test in \cref{sec:testing}.
We give a proof-of-concept formalization of a more sophisticated, heap-like sharing model
based on pointer provenance in \cref{app:ptr}\bcp{first
  mention of this appendix; if it's important, mention it in the
  outline in section I.  Otherwise drop it.}, represented by the dark gray
operations, but we do not test this.

\bcp{As a reader, I'm feeling a bit overwhelmed at this point.
  Everything feels fluid and flexible---''It could be like this, but
  it could also be another way if you want...''  I know we're past the
  point of doing more than tweaking the way we talk about things, but
  if there are any tweaks that reduce the sense of flexibility, let's
  do them!}

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l | l |}
    \hline
    \(\psi \in \Psi\)\bcp{huh?} & Parameters \\
    \hline
    \example \(\mathbf{call}\) & target address, argument registers \\
    \testing & stack arguments (base, offset \& size) \\
    \example \(\mathbf{return}\) & \\
    \example \(\mathbf{alloc}\) & offset \& size \\
    \testing & public flag \\
    \example \(\mathbf{dealloc}\) & offset \& size \\
    \testing \(\mathbf{tail call}\) & (same as for \(\mathbf{call}\)) \\
    \hline
    \multicolumn{2}{|c|}{{\it Pointer provenance operations}} \\
    \hline
    \theory \(\mathbf{promote}\) & register, offset \& size \\
    \theory \(\mathbf{propagate}\) & source register/address \\
    \theory & destination register/address \\
    \theory \(\mathbf{clear}\) & target register/address \\
    \hline
  \end{tabular}
\end{center}
\caption{Security-relevant operations and their parameters.}
  \label{tab:psi}
\end{table}
